# Correlation and regression

```{r setup, include=FALSE}
source("global_parameters.R")
knitr::opts_chunk$set(fig.width=fig.width, fig.height=fig.height)
```

The function `cor` can be used to compute Pearson's $r$ correlation coefficient. For example, using the `iris` dataset we can compute the correlation between petal length and petal width for the setosa iris species as follows:
```{r}
data(iris)
iris_setosa = iris[iris$Species == "setosa",]
cor(iris_setosa$Petal.Length, iris_setosa$Petal.Width)
```

Kendall's $\tau$, or Spearman's $\rho$ rank correlation coefficients can be obtained with the same function by changing the `method` argument from its default value (`pearson`) to `kendall`, or `spearman`, e.g.:
```{r}
cor(iris_setosa$Petal.Length, iris_setosa$Petal.Width, method="kendall")
```

A significance test on the correlation coefficient can be obtain with the `cor.test` function:
```{r}
cor.test(iris_setosa$Petal.Length, iris_setosa$Petal.Width)
```

If we pass a matrix or dataframe to the `cor` function, it will compute the correlations between each pair of columns of the matrix/dataframe:
```{r}
cor(iris_setosa[, 1:4])
```

in the code above we selected columns 1-4 to remove the non-numerical column specifying the iris species. The `cor.test` function, however, cannot be passed a matrix or dataframe as input. The `rcorr` function in the [`Hmisc`](https://cran.r-project.org/web/packages/Hmisc/index.html) package can be used to compute correlations and significance tests between all the columns of a matrix:
```{r}
corr_mat = Hmisc::rcorr(as.matrix(iris_setosa[,1:4]))
corr_mat$r # correlation coefficients
corr_mat$P # corresponding p-values
```

note that `rcorr` can compute only Pearson's $r$ or Spearman's $\rho$ correlation coefficients, by passing an argument named `type` (rather than `method`, as for the `cor` function).

Functions to compute partial, and semi-partial correlations are available in the [`ppcor`](https://cran.r-project.org/web/packages/ppcor/) package.

## Linear regression

Regression models are expressed in R with a symbolic syntax that is clean and convenient. If you have a dependent variable `rts` and you want to check the influence of an independent variable `age` on it, this would be expressed as:
```{r, eval=FALSE}
rtfit = lm(rts ~ age)
plot(rts ~ age)
abline(rtfit)
```
if you want to add another predictor `score`
```{r, eval=FALSE}
rtfit2 = lm(rts ~ age + score)
```  


