# Plotly {#plotly}

The ["plotly for R"](https://cpsievert.github.io/plotly_book/) book [@Sievert2017] is the best introduction for
learning plotly.


```{r plotly1, message=FALSE, warning=FALSE, fig.cap="Plotly example"}
library(plotly)
x = rnorm(10); y=rnorm(10)
plot_ly(x=x, y=y, type="scatter", mode="markers")
```

```{r plotly2}
plot_ly(x=x, y=y, type="scatter", mode="markers",
        marker=list(color="black" , size=15 , opacity=0.8))
```

## Using plotly with knitr

Plotly figures are rendered directly in the html output with knitr:

````markdown
`r ''````{r chunk-label, fig.cap = 'A figure caption.'}
plot_ly(economics, x = ~date, y = ~unemploy / pop)
```
````

please note that if the plot is assigned to a variable you need to call that variable in the code chunk for the plot to be rendered:


````markdown
`r ''````{r chunk-label, fig.cap = 'A figure caption.'}
p = plot_ly(economics, x = ~date, y = ~unemploy / pop)
p #call the variable storing the plot to render it
```
````

Plotly figures can also appear in pdf files generate by knitr if the `webshot` package is installed. Besides this package, you will also need to have PhantomJS (http://phantomjs.org/) installed on your system. You can install both `webshot` and PhantomJS from within R with the following commands:

```{r eval=FALSE}
install.packages("webshot")
webshot::install_phantomjs()
```


